\chapter{Advanced usage}
\label{chap:Advanced usage}

\section{Cloning objects}
\label{sec:Cloning objects}

The \cppinline{Sampler<T>} and \cppinline{Particle<T>} objects have copy
constructors, assignment operators, move constructors, and move assignment
operators that behaves exactly the way as \cpp programmers would expect.
However, these behaviors are not always desired. For example, in
\textcite{stpf} a stable particle filter in high-dimensions was developed.
Without going into the details, the algorithm consists of a particle system
where each particle is itself a particle filter. And thus when resampling the
global system, the \cppinline{Sampler<T>} object will be copied, together with
all of its sub-objects. This include the \rng system within the
\cppinline{Particle<T>} object. Even if the user does not use this \rng system
for random number generating within user defined operations, one of these \rng
will be used for resampling by the \cppinline{Particle<T>} object. Direct
copying the \cppinline{Sampler<T>} object will lead to multiple local filters
start to generating exactly the same random numbers in the next iteration. This
is an undesired side effects. In this situation, one can clone the sampler with
the following method,
\begin{cppcode}
  auto new_sampler = sampler.clone(new_rng);
\end{cppcode}
where \cppinline{new_rng} is a boolean value. If it is \cppinline{true}, then
an exact copy of \cppinline{sampler} will be returned, except it will have the
\rng system re-seeded. If it is \cppinline{false}, then the above assignemnt
behaves exactly the same as
\begin{cppcode}
  auto new_sampler = sampler;
\end{cppcode}
Alternatively, the contents of an existing \cppinline{Sampler<T>} object can be
replaced from another one by the following method,
\begin{cppcode}
  sampler.clone(other_sampler, retain_rng);
\end{cppcode}
where \cppinline{retain_rng} is a boolean value. If it is \cppinline{true},
then the \rng system of \cppinline{other_sampler} is not copied and its own
\rng system is retained. If it is \cppinline{false}, then the above call
behaves exactly the same as
\begin{cppcode}
  sampler = other_sampler;
\end{cppcode}
The above method also supports move semantics. Similar \cppinline{clone}
methods exist for the \cppinline{Particle<T>} class.

\section{Customizing member types}
\label{sec:Customizing member types}

The \cppinline{Particle<T>} class has a few member types that can be replaced
by the user. If the class \cppinline{T} has the corresponding types, then the
member type of \cppinline{Particle<T>} will be replaced. For example, given the
following declarations inside class \cppinline{T},
\begin{cppcode}
  class T
  {
      public:
      using size_type = int;
      using weight_type = /* User defined type */;
      using rng_set_type = RNGSetTBB<AES256_4x32>;
  };
\end{cppcode}
The corresponding \cppinline{Particle<T>::size_type}, etc., will have their
defaults replaced with the above types.

A note on \cppinline{weight_type}, it needs to provide the following method,
\begin{cppcode*}{texcomments}
  w.ess();           // Get $\text{\normalfont\textsc{ess}}$
  w.set_equal();     // Set $W^{(i)} = 1/N$
  w.resample_size(); // Get the size $N$.
  w.resample_data(); // Get a pointer to normalized weights
\end{cppcode*}
For the library's default class \cppinline{Weight}, the last two calls are the
same as \cppinline{w.size()} and \cppinline{w.data()}. However, this does not
need to be so. For example, below is the outline of an implementation of
\cppinline{weight_type} for distributed systems, assuming each computing node
has been allocated $N_r$ particles.
\begin{cppcode*}{texcomments}
  class WeightMPI
  {
      public:
      double ess()
      {
          double local = /* $\sum_{i=1}^{N_r}(W^{(i)})^2$ */;
          double global = /* Gather local from each all nodes */;
          // Broadcast the value of global

          return 1 / global;
      }

      std::size_t resample_size() { return /* $\sum N_r$ */; }

      double *resample_data()
      {
          if (rank == 0) {
              // Gather all normalized weights into a member data on this node
              // Say resample\_weight\_
              return resample_weight_.data();
          } else {
              return nullptr;
          }
      }

      void set_equal()
      {
          // Set all weights to $1 / \sum N_r$
          // Synchronization
      }

      void set(const double *v)
      {
          // Set $W^{(i)} = v_i$ for $i = 1,\dots,N_r$
          // Compute $S_r = \sum_{i=1}^{N_r} W^{(i)}$
          // Gathering $S_r$, compute $S = \sum S_r$
          // Broadcast $S$
          // Set $W^{(i)} = W^{(i)} / S$ for $i = 1,\dots,N_r$
      }
  };
\end{cppcode*}
When \cppinline{Particle<T>} performs resampling, it checks if the pointer
returned by \cppinline{w.resample_data()} is a null pointer. It will only
generate the vector $\{a_i\}_{i=1}^N$ (see section~\ref{sub:State}) when it is
not a null pointer. And then a pointer to this vector is passed to
\cppinline{T::copy}. Of course, the class \cppinline{T} also needs to provide a
suitable method \cppinline{copy} that can handle the distributed system. By
defining suitable \cppinline{WeightMPI} and \cppinline{T::copy}, the library
can be extended to handle distributed systems.

\section{Extending \protect\spt}
\label{sec:Extending SP}

The \cppinline{SingleParticle<T>} can also be extended by the user. We have
already see in section~\ref{sub:State} that if class \cppinline{T} is a
subclass of \cppinline{StateMatrix}, \cppinline{SingleParticle<T>} can have
additional methods to access the state. This class can be extended by defining
a member class template inside class \cppinline{T}. For example, for the simple
particle filter in section~\ref{sec:A simple particle filter}, we can redefine
the \cppinline{PFState} as the following,
\begin{cppcode}
  using PFStateBase = StateMatrix<RowMajor, 4, double>;

  template <typename T>
  using PFStateSPBase = PFStateBase::single_particle_type<T>;

  class PFState : public PFStateBase
  {
      public:
      using PFStateBase::StateMatrix;

      template <typename S>
      class single_particle_type : public PFStateSPBase<S>
      {
          public:
          using PFStateSPBase<S>::single_particle_type;

          double &pos_x() { return this->state(0); }
          double &pos_y() { return this->state(1); }
          double &vel_x() { return this->state(2); }
          double &vel_y() { return this->state(3); }

          // Return $\ell(X_t^{(i)}|Y_t)$
          double log_likelihood(std::size_t t);
      };

      void read_data(const char *param);

      private:
      Vector<double> obs_x_;
      Vector<double> obs_y_;
  };
\end{cppcode}
And later, we can use these methods when implement \cppinline{PFInit} etc.,
\begin{cppcode}
  class PFInit : public InitializeTBB<PFState, PFInit>
  {
      public:
      void eval_param(Particle<PFState> &particle, void *param);

      void eval_pre(Particle<PFState> &particle);

      std::size_t eval_sp(SingleParticle<PFState> sp)
      {
          NormalDistribution<double> norm_pos(0, 2);
          NormalDistribution<double> norm_vel(0, 1);
          sp.pos_x() = norm_pos(sp.rng());
          sp.pos_y() = norm_pos(sp.rng());
          sp.vel_x() = norm_vel(sp.rng());
          sp.vel_y() = norm_vel(sp.rng());
          w_[sp.id()] = sp.log_likelihood(0);

          return 0;
      }

      void eval_post(Particle<PFState> &particle);

      private:
      Vector<double> w_;
  };
\end{cppcode}
It shall be noted that, it is important to keep
\cppinline{single_particle_type} small and copying the object efficient. The
library will frequently pass argument of \cppinline{SingleParticle<T>} type by
value.

\subsection{Compared to custom state type}
\label{sub:Compared to custom state type}

One can also write a custom state type. For example,
\begin{cppcode}
  class PFStateSP
  {
      public:
      double &pos_x() { return pos_x_; }
      double &pos_y() { return pos_y_; }
      double &vel_x() { return vel_x_; }
      double &vel_y() { return vel_y_; }

      double log_likelihood(double obs_x, double obs_y) const;

      private:
      double pos_x_;
      double pos_y_;
      double vel_x_;
      double vel_y_;
  };
\end{cppcode}
And the \cppinline{PFState} class will be defined as,
\begin{cppcode}
  using PFStateBase = StateMatrix<RowMajor, 1, PFStateSP>;

  class PFState : public PFStateBase
  {
      public:
      using PFStateBase::StateMatrix;

      double log_likelihood(std::size_t t, std::size_t i) const
      {
          return this->state(i, 0).log_likelihood(obs_x_[t], obs_y_[t]);
      }

      void read_data(const char *param);

      private:
      Vector<double> obs_x_;
      Vector<double> obs_y_;
  };
\end{cppcode}
The implementation of \cppinline{PFInit}, etc., will be similar. Compared to
extending the \cppinline{SingleParticle<T>} type, this method is perhaps more
intuitive. Functionality-wise, they are almost identical. However, there are a
few advantages of extending \cppinline{SingleParticle<T>}. First, it allows
more compact data storage. Consider a situation where the state space is best
represented by a real and an integer. The most intuitive way might be the
following,
\begin{cppcode}
  class S
  {
      public:
      double &x() { return x_; }
      double &u() { return u_; }

      private:
      double x_;
      int u_;
  };

  class T : StateMatrix<RowMajor, 1, S>;
\end{cppcode}
However, the type \cppinline{S} will need to satisfy the alignment requirement
of \cppinline{double}, which is 8-bytes on most platforms. However, its size
might not be a multiple of 8-bytes. Therefore the type will be padded and the
storage of a vector of such type will not be as compact as possible. This can
affect performance in some situations. An alternative approach would be the
following,
\begin{cppcode}
  class T
  {
      public:
      template <typename S>
      class single_particle_type : SingleParticleBase<S>
      {
          public:
          using SingleParticleBase<S>::SingleParticleBase;

          double &x() { return this->particle().x_[this->id()]; }
          double &u() { return this->particle().u_[this->id()]; }
      };

      private:
      Vector<double> x_;
      Vector<int> u_;
  };
\end{cppcode}
By extending \cppinline{SingleParticle<T>}, it provides the same easy access to
each particle. However, now the state values are stored as two compact vectors.

A second advantage is that it allows easier access to the raw data. Consider
the implementation \cppinline{PFMEval} in section~\ref{sub:Implementations}.
It is rather redundant to copy each value of the two positions, just so later
we can compute weighted sums from them. Recall that in
section~\ref{sub:Monitor} we showed that a monitor that compute the final
results directly can also be added to a sampler. Therefore, we might
implement \cppinline{PFMEval} as the following,
\begin{cppcode*}{texcomments}
  class PFMEval
  {
      public:
      void operator()(std::size_t t, std::size_t dim,
          Particle<PFState> &particle, double *r)
      {
          cblas_dgemv(CblasRowMajor, CblasTrans, particle.size(), dim, 1,
              particle.value().data(), particle.value().dim(),
              particle.weight().data(), 1, 0, r, 1);
      }
  };
\end{cppcode*}
And it can be added to a sampler as,
\begin{cppcode}
  sampler.monitor("pos", 2, PFMEval(), true);
\end{cppcode}
For this particular case, the performance benefit is small. But the possibility
of accessing compact vector as raw data allows easier interfacing with external
numerical libraries. If we implemented \cppinline{PFState} with the alternative
approach shown earlier, the above direct invoking of \cppinline{cblas_dgemv}
will not be possible.
