\chapter{Random number generating}
\label{chap:Random number generating}

The library has a comprehensive \rng system to facilitate implementation of
Monte Carlo algorithms. Similar to the standard library header \verb|<random>|,
there are mainly two parts of this system. The first is a set of \rng{} engines
that generate random integers. The other is a set of distribution generators.
The former are documented in sections~\ref{sec:Counter-based RNG}
to~\ref{sec:MKL RNG}, and the later in section~\ref{sec:Distributions}. Apart
from these, the library also provides facilities for vectorized random number
generating (section~\ref{sec:Vectorized random number generating}) and using
multiple \rng{}s in parallel programs (section~\ref{sec:Multiple RNG streams}).

\section{Vectorized random number generating}
\label{sec:Vectorized random number generating}

Before we discuss other features, we first introduce a generic function
\verb|rand|, which provides vectorized random number generating. There are two
versions. The first operates on \rng engines and generates random integers,
\begin{Verbatim}
  template <typename RNGType>
  inline void rand(
      RNGType &rng, std::size_t n, typename RNGType::result_type *r);
\end{Verbatim}
The effect of the function call,
\begin{Verbatim}
  rand(rng, n, r);
\end{Verbatim}
is equivalent to the loop,
\begin{Verbatim}
  for (std::size_t i = 0; i != n; ++i)
      r[i] = rng();
\end{Verbatim}
The results will always be the same unless a non-deterministic \rng is used.
For some \rng{}s implemented in the library, the vectorized version may have
considerable performance advantage.

The second version of \verb|rand| is for generating distribution random
numbers,
\begin{Verbatim}
  template <typename RNGType, typename DistributionType>
  inline void rand(RNGType &rng, const DistributionType &distribution,
      std::size_t n, typename DistributionType::result_type *r);
\end{Verbatim}
For example,
\begin{Verbatim}
  NormalDistribution<double> normal;
  rand(rng, normal, n, r);
\end{Verbatim}
This is similar to the following loop,
\begin{Verbatim}
  for (std::size_t i = 0; i != n; ++i)
      r[i] = normal(rng);
\end{Verbatim}
Depending on the type of \verb|rng| and the distribution (including its
parameters), the vectorized version may have superior performance. However, the
results will not be exactly the same as using a loop.

\subsection{Performance measurement}
\label{sub:Performance measurement}

All performance results shown later in this chapter is measured in single core
cycles per bytes (cpB) for \rng{}s, or cycles per element (cpE) for
distributions (\verb|double| precision). The processor used to measure the
performance is an Intel Core i7-4960\textsc{hg} \cpu. Three compilers are
tested. The \llvm clang (version~3.8), the \gnu{} \gcc (version~6.1), and the
Intel \cpp compiler (version~2016 update~3). When multiple compilers are
tested, they are labeled ``\llvm'', ``\gnu'', and ``Intel'', respectively in
tables. If only results from one compiler are shown, then unless stated
otherwise, it is the \llvm clang compiler. The operating system is Mac OS X
(version~11.1). Depending on the \cpu, the compiler, and the operating system,
performance may vary considerably. The library does not attempt to optimize for
any particular platform. However, given accelerated vectorized mathematical
functions (seed section~\ref{sec:Vectorized functions}), for any given
platform, the relative performance advantage to alternatives, such as the
standard library, is usually substantial.

For \rng engines, we measure the performance of generating random bits instead
of the raw output from the engines. All internal usages of \rng{}s in the
library first transfer the raw output to unsigned integers uniform on the set
$\{0,\dots,2^W-1\}$, $W\in\{32,64\}$ (see section~\ref{sub:Uniform bits
  distribution}). Direct use of the raw output from \rng engines is rare in
applications.

Two usage cases are considered. The first is the performance of generating
random integers one by one,
\begin{Verbatim}
  UniformBitsDistribution<std::uint64_t> rbits;
  for (std:size_t i = 0; i != n; ++i)
      r[i] = rbits(rng);
\end{Verbatim}
The second is the vectorized performance,
\begin{Verbatim}
  rand(rng, rbits, n, r.data());
\end{Verbatim}
In both cases, we repeat the simulations 100 times, each time with the number
of elements $n$ chosen randomly between 5,000 and 10,000. The total number of
cycles of the 100 simulations are recorded, and then divided by the total
number of bytes generated. This gives the performance measurement in cpB. This
experiment is repeated ten times, and the best results are shown. The two cases
are labeled ``Loop'' and ``\verb|rand|'', respectively.

For the performance of distributions, we measure four methods of drawing random
numbers from the same distribution. The procedures is similar, except now we
measure the performance in cpE. First, if the distribution is available in the
standard library or the Boost\footnote{\url{http://www.boost.org}} library, we
measure the following case,
\begin{Verbatim}
  std::normal_distribution<double> rnorm_std(0, 1);
  for (std::size_t i = 0; i = n; ++i)
      r[i] = rnorm_std(rng);
\end{Verbatim}
Second, we measure the performance of the library's implementation,
\begin{Verbatim}
  NormalDistribution<double> rnorm_vsmc(0, 1);
  for (std::size_t i = 0; i = n; ++i)
      r[i] = rnorm_vsmc(rng);
\end{Verbatim}
The third is the vectorized performance,
\begin{Verbatim}
  rand(rng, rnorm_vsmc, n, r.data());
\end{Verbatim}
For all the three above, the \rng is \verb|ARSx8| (section~\ref{sub:AES-NI
  instructions based RNG}). The last is when \rng is \verb|MKL_SFMT19937|
(section~\ref{sec:MKL RNG}),
\begin{Verbatim}
  MKL_SFMT19937 rng_mkl;
  rand(rng_mkl, rnorm_vsmc, n, r.data());
\end{Verbatim}
In this case, not only the \rng itself is faster, the distribution might also
use \mkl routines. The four cases are labeled ``\std/Boost'', ``\vsmc'',
``\verb|rand|'' and ``\mkl'', respectively.

\section{Counter-based RNG}
\label{sec:Counter-based RNG}

The standard library provides a set of \rng engines (performance data in
Table~\ref{tab:Performance of standard library RNG}). Unfortunately, none of
them are suitable for parallel computing without considerable efforts. To
illustrate the problem, consider the situation where there are two threads that
need to generate random numbers. And thus two \rng engine instances need to be
created for each thread. Let the random numbers generated by them be
$\{r_i^1\}_{i>0}$ and $\{r_i^2\}_{i>0}$. Because of the deterministic and
recursive natural of the these \rng{}s, there exists $k\in\Integer$, such that,
$r_i^1 = r_{i + k}^2$ for all $i > \max\{0, -k\}$. If $\Abs{k}$ is larger than
or close to the total number of random numbers required in each thread, then
this situation might not be an issue. However, there is no easy way to ensure
such a condition.

There are two standard solutions to this problem. The first is to use
sub-streams for each thread. For example,
\begin{Verbatim}
  std::mt19937 rng;

  // Thread k
  std::mt19937 rng_k = rng;
  rng_k.discard(n * k);
\end{Verbatim}
where $n$ is the number of random numbers required by each thread. The $k$\ith
thread only uses the random numbers in the sub-stream $\{r_i\}_{nk < i \le
  n(k+1)}$. For this to work, the \rng needs a fast \verb|discard|
implementation, preferably with $\calO(1)$ cost. In addition, one needs to
manage \rng{}s explicitly for each thread, which prevents this method to be
used in environments where threads are created implicitly, such as using \tbb
for parallelization.

The second solution is to use a leap-frog algorithm, such that each thread will
use the elements $\{r_{iK + k}\}_{i>0}$ of the stream, where $K$ is the total
number of threads. This is not directly supported in the standard library, but
can be emulated,
\begin{Verbatim}
  std::mt19937 rng;

  // Thread k
  std::mt19937 tmp = rng;
  tmp.discard(k);
  std::discard_block_engine<std::mt19937, K, 1> rng_k(tmp);
\end{Verbatim}
This not only requires managing the \rng{}s explicitly, but also knowing the
number of threads at compile-time, which prevents it to be used in any
applications using dynamic parallelization. And similar to the first solution,
it cannot be used when threads are created implicitly.

\begin{table}
  \input{tab/rng_engine_std}%
  \caption{Performance of standard library \protect\rng}
  \label{tab:Performance of standard library RNG}
\end{table}

The development by \textcite{Salmon:2011um} made high performance parallel \rng
much more accessible. The \rng{}s introduced in the paper use bijection $f_k$,
such that, for a sequence $\{c_i = i\}_{i\ge0}$, the sequence $\{y_i =
f_k(c_i)\}_{i\ge0}$ appears random. In addition, for $k_1 \ne k_2$, $f_{k_1}$
and $f_{k_2}$ will generate two sequences that appear statistically
independent. Compared to more conventional \rng{}s which use recursions $y_i =
f_k(y_{i - 1})$, these counter-based \rng{}s are much easier to setup in a
parallelized environment. If $c$, the counter, is an unsigned integer with $b$
bits, and $k$, the key, is an unsigned integer with $d$ bits. Then for each
$k$, the \rng has a period $2^b$. And there can be at most $2^d$ independent
streams. Another way is to view this kind of \rng{}s is that, it is one \rng{}
with state $\{c, k\}$ for $0 \le c < 2^b$ and $0 \le k < 2^d$. And thus it has
a period $2^{b + d}$. And the division of the state into counter and key
provides an easy way to setup a large number of sub-streams.

Of course, not any sequence of counters and keys are suitable. The \rng{}s are
still deterministic. Since $f_k$ for any given $k$ is a bijection, the sequence
of counter $\{c_i = f_k^{-1}(i)\}_{i\ge0}$ will of course produce a very
regular sequence $\{y_i = i\}_{i\ge0}$. However, for regular sequences, such
as $\{c_i = i\}_{i\ge0}$ and $\{k_j = j\}_{j\ge0}$, the resulting sequences
$\{\{y_i\}_{i\ge0}^j\}_{j\ge0}$ appear random. See \textcite{Salmon:2011um} for
more details.

Table~\ref{tab:Counter-based RNG} lists all counter-based \rng{}s implemented
in the library, along with the bits of the counter and the key. They all
output 32-bits unsigned integers uniform on the set $\{0,\dots,2^{32}-1\}$. For
64-bits output, a suffix \verb|_64| may be appended to the corresponding \rng
engine names. For example, \verb|Threefry4x64| and \verb|Threefry4x64_64| both
generate the same 256-bits random integers internally. The only difference is
that \verb|operator()| of the former returns 32 or those 256 bits each time it
is executed, while the later returns 64 bits.

All \rng{}s in Table~\ref{tab:Counter-based RNG} are actually type aliases.
More generally the library defines the following class template as the
interface,
\begin{Verbatim}
  template <typename ResultType, typename Generator>
  class CounterEngine;
\end{Verbatim}
where \verb|ResultType| shall be an unsigned integer type and \verb|Generator|
is the class that actually implements the algorithm. See the reference manual
for details of the generator type. For most users, those implemented in the
library are sufficient. They are introduced in the next few sections. A few
configuration macros of these generators are listed in
Table~\ref{tab:Configuration macros for counter-based RNG} and will be referred
to later.

\begin{table}
  \tbfigures
  \begin{tabularx}{\textwidth}{p{3in}LL}
    \toprule
    Class & Counter bits & Key bits \\
    \midrule
    \verb|AES128x1|, \verb|ARS128x2|, \verb|AES128x4|, \verb|AES128x8|
    & 128 & 128 \\
    \verb|AES192x1|, \verb|ARS192x2|, \verb|AES192x4|, \verb|AES192x8|
    & 128 & 192 \\
    \verb|AES256x1|, \verb|AES256x2|, \verb|AES256x4|, \verb|AES256x8|
    & 128 & 256 \\
    \verb|ARSx1|, \verb|ARSx2|, \verb|ARSx4|, \verb|ARSx8| & 128 & 128 \\
    \verb|Philox2x32|    & 64   & 32   \\
    \verb|Philox2x64|    & 128  & 64   \\
    \verb|Philox4x32|    & 128  & 64   \\
    \verb|Philox4x64|    & 256  & 128  \\
    \verb|Threefry2x32|  & 64   & 64   \\
    \verb|Threefry2x64|  & 128  & 128  \\
    \verb|Threefry4x32|  & 128  & 128  \\
    \verb|Threefry4x64|  & 256  & 256  \\
    \verb|Threefry8x64|  & 512  & 512  \\
    \verb|Threefry16x64| & 1024 & 1024 \\
    \bottomrule
  \end{tabularx}
  \caption{Counter-based \protect\rng}
  \label{tab:Counter-based RNG}
\end{table}

\begin{table}
  \begin{tabularx}{\textwidth}{LL}
    \toprule
    Macro & Default \\
    \midrule
    \verb|VSMC_RNG_AES128_ROUNDS|          & \verb|10| \\
    \verb|VSMC_RNG_AES192_ROUNDS|          & \verb|12| \\
    \verb|VSMC_RNG_AES256_ROUNDS|          & \verb|14| \\
    \verb|VSMC_RNG_ARS_ROUNDS|             & \verb|5|  \\
    \verb|VSMC_RNG_AES_NI_BLOCKS|          & \verb|8|  \\
    \verb|VSMC_RNG_PHILOX_ROUNDS|          & \verb|10| \\
    \verb|VSMC_RNG_PHILOX_VECTOR_LENGTH|   & \verb|4|  \\
    \verb|VSMC_RNG_THREEFRY_ROUNDS|        & \verb|20| \\
    \verb|VSMC_RNG_THREEFRY_VECTOR_LENGTH| & \verb|4|  \\
    \bottomrule
  \end{tabularx}
  \caption{Configuration macros for counter-based \protect\rng}
  \label{tab:Configuration macros for counter-based RNG}
\end{table}

\subsection{\protect\aesni instructions based \protect\rng}
\label{sub:AES-NI instructions based RNG}

The \aesni\footnote{\url{https://en.wikipedia.org/wiki/AES_instruction_set}}
instructions based \rng{}s in \textcite{Salmon:2011um} are implemented in the
following generator,
\begin{Verbatim}
  template <typename KeySeqType, std::size_t Rounds, std::size_t Blocks>
  class AESNIGenerator;
\end{Verbatim}
The corresponding \rng engine is,
\begin{Verbatim}
  template <typename ResultType, typename KeySeqType, std::size_t Rounds,
      std::size_t Blocks>
  using AESNIEngine =
      CounterEngine<ResultType, AESNIGenerator<KeySeqType, Rounds, Blocks>>;
\end{Verbatim}
where \verb|KeySeqType| is the class used to generate the sequences of round
keys. The parameter \verb|Rounds| is the number of rounds of \aes encryption to
be performed. See the reference manual for details of how to define the key
sequence class. The \aesni encryption instructions have a latency of seven or
eight cycles, while they can be issued at every cycle. Therefore better
performance can be achieved if multiple 128-bits random integers are generated
at the same time. This is specified by the template parameter \verb|Blocks|.
Larger blocks, up to eight, might improve performance. But this is at the cost
of larger state size. Without going into details, there are four types of
sequence of round keys implemented by the library,
\begin{Verbatim}
  template <std::size_t Rounds>
  using AES128KeySeq =
      internal::AESKeySeq<Rounds, internal::AES128KeySeqGenerator>;

  template <std::size_t Rounds>
  using AES192KeySeq =
      internal::AESKeySeq<Rounds, internal::AES192KeySeqGenerator>;

  template <std::size_t Rounds>
  using AES256KeySeq =
      internal::AESKeySeq<Rounds, internal::AES256KeySeqGenerator>;

  template <typename Constants = ARSConstants>
  using ARSKeySeq = internal::ARSKeySeqImpl<Constants>;
\end{Verbatim}
and correspondingly four \rng engines,
\begin{Verbatim}
  template <typename ResultType, std::size_t Rounds = VSMC_RNG_AES128_ROUNDS,
      std::size_t Blocks = VSMC_RNG_AES_NI_BLOCKS>
  using AES128Engine =
      AESNIEngine<ResultType, AES128KeySeq<Rounds>, Rounds, Blocks>;

  template <typename ResultType, std::size_t Rounds = VSMC_RNG_AES192_ROUNDS,
      std::size_t Blocks = VSMC_RNG_AES_NI_BLOCKS>
  using AES192Engine =
      AESNIEngine<ResultType, AES192KeySeq<Rounds>, Rounds, Blocks>;

  template <typename ResultType, std::size_t Rounds = VSMC_RNG_AES256_ROUNDS,
      std::size_t Blocks = VSMC_RNG_AES_NI_BLOCKS>
  using AES256Engine =
      AESNIEngine<ResultType, AES256KeySeq<Rounds>, Rounds, Blocks>;

  template <typename ResultType, std::size_t Rounds = VSMC_RNG_ARS_ROUNDS,
      std::size_t Blocks = VSMC_RNG_AES_NI_BLOCKS,
      typename Constants = ARSConstants>
  using ARSEngine =
      AESNIEngine<ResultType, ARSKeySeq<Constants>, Rounds, Blocks>;
\end{Verbatim}
The first three are equivalent to \aes-128, \aes-192 and \aes-256 block ciphers
used in counter mode. The last is the \ars algorithm introduced by
\textcite{Salmon:2011um}. The last template parameter \verb|Constants| of
\verb|ARSKeySeq| and \verb|ARSEngine| is a trait class that defines the
constants of the Weyl's sequence. See \textcite{Salmon:2011um} for details. The
defaults are taken from the paper. To use an alternative pair of 64-bits
integers as the constants, one can define and use a trait class as the
following,
\begin{Verbatim}
  template <std::size_t>
  struct NewWeylConstant;

  template<>
  struct NewWeylConstant<0>
  {
      static constexpr std::uint64_t value = FIRST_CONSTANT;
  };

  template<>
  struct NewWeylConstant<1>
  {
      static constexpr std::uint64_t value = SECOND_CONSTANT;
  };

  struct NewConstants
  {
      template <std::size_t I>
      using weyl = NewWeylConstant<I>;
  };

  using NewARS = ARSEngine<ResultType, Rounds, NewConstants>;
\end{Verbatim}
Alternative methods are also possible. The only requirement is that, the
following statement,
\begin{Verbatim}
  template <std::size_t I>
  using weyl = typename Constants::template weyl<I>;
\end{Verbatim}
shall define the type \verb|weyl| such that it has a static constant expression
member data \verb|value| that is the \verb|I|\ith Weyl constant. A few type
aliases are defined for convenience. For example,
\begin{Verbatim}
  using ARSx8    = ARSEngine<std::uint32_t, VSMC_RNG_ARS_ROUNDS, 8>;
  using ARSx8_64 = ARSEngine<std::uint64_t, VSMC_RNG_ARS_ROUNDS, 8>;
  using ARS      = ARSEngine<std::uint32_t>;
  using ARS_64   = ARSEngine<std::uint64_t>;
\end{Verbatim}
The engine \verb|ARS| is the library's default \rng if \aesni instructions are
supported. Aliases for block sizes 1, 2, 4 and 8 are defined for all four
algorithms, as well as both 32- and 64-bits output versions. These aliases are
listed in Table~\ref{tab:Counter-based RNG}. The performance of these engines
depends on a few factors, such as \cpu types, compilers, operating systems,
etc. See Tables~\ref{tab:Performance of AES128Engine} to~\ref{tab:Performance
  of ARSEngine} for performance data. In any case, the performance is good
enough even for the most demanding applications. The library does not attempt
to optimize the algorithm for any particular platform. In realistic
applications, the performance of \rng is unlikely to become a bottle neck. Note
that, the best performance is obtained with the vectorized \verb|rand| function
(see section~\ref{sec:Vectorized random number generating}).

\begin{table}
  \input{tab/rng_engine_aes128}%
  \caption{Performance of \protect\texttt{AES128Engine}}
  \label{tab:Performance of AES128Engine}
\end{table}

\begin{table}
  \input{tab/rng_engine_aes192}%
  \caption{Performance of \protect\texttt{AES192Engine}}
  \label{tab:Performance of AES192Engine}
\end{table}

\begin{table}
  \input{tab/rng_engine_aes256}%
  \caption{Performance of \protect\texttt{AES256Engine}}
  \label{tab:Performance of AES256Engine}
\end{table}

\begin{table}
  \input{tab/rng_engine_ars}%
  \caption{Performance of \protect\texttt{ARSEngine}}
  \label{tab:Performance of ARSEngine}
\end{table}

\subsection{Philox}
\label{sub:Philox}

The Philox algorithm in \textcite{Salmon:2011um} is implemented in the
following generator,
\begin{Verbatim}
  template <typename T, std::size_t K = VSMC_RNG_PHILOX_VECTOR_LENGTH,
      std::size_t Rounds = VSMC_RNG_PHILOX_ROUNDS,
      typename Constants = PhiloxConstants<T, K>>
  class PhiloxGenerator;
\end{Verbatim}
The corresponding \rng engine is,
\begin{Verbatim}
  template <typename ResultType, typename T = ResultType,
      std::size_t K = VSMC_RNG_PHILOX_VECTOR_LENGTH,
      std::size_t Rounds = VSMC_RNG_PHILOX_ROUNDS,
      typename Constants = PhiloxConstants<T, K>>
  using PhiloxEngine =
      CounterEngine<ResultType, PhiloxGenerator<T, K, Rounds, Constants>>;
\end{Verbatim}
The default vector length and the number of rounds can be changed by
configuration macros listed in Table~\ref{tab:Configuration macros for
  counter-based RNG}. There is no limit on the template parameter \verb|K| or
\verb|Rounds|, nor any limitation on \verb|T| except that it has to be an
unsigned integer type. See \textcite{Salmon:2011um} on the most general form of
the algorithm. However, the library only provides default constants for 32- and
64-bits unsigned integer type \verb|T| and \verb|K| taking the values \verb|2|
or \verb|4|. These four engines are defined as type aliases for convenience,
\begin{Verbatim}
  template <typename ResultType>
  using Philox2x32Engine = PhiloxEngine<ResultType, std::uint32_t, 2>;

  template <typename ResultType>
  using Philox4x32Engine = PhiloxEngine<ResultType, std::uint32_t, 4>;

  template <typename ResultType>
  using Philox2x64Engine = PhiloxEngine<ResultType, std::uint64_t, 2>;

  template <typename ResultType>
  using Philox4x64Engine = PhiloxEngine<ResultType, std::uint64_t, 4>;
\end{Verbatim}
Type aliases for 32- and 64-bits \verb|ResultType| are also defined, as listed
in Table~\ref{tab:Counter-based RNG}. To use the engine with \verb|K| taking
values larger than four, or \verb|T| being unsigned integer type with bits
other than 32 or 64, one needs to provide a suitable trait class,
\verb|Constant|. It is similar to that of \verb|ARSEngine|. See the reference
manual of \verb|PhiloxConstants| for an example of how to define it. The
performance data is in Table~\ref{tab:Performance of PhiloxEngine}.

\begin{table}
  \input{tab/rng_engine_philox}%
  \caption{Performance of \protect\texttt{PhiloxEngine}}
  \label{tab:Performance of PhiloxEngine}
\end{table}

\subsection{Threefry}
\label{sub:Threefry}

The Threefry algorithm in \textcite{Salmon:2011um} is implemented in the
following generator,
\begin{Verbatim}
  template <typename T, std::size_t K = VSMC_RNG_THREEFRY_VECTOR_LENGTH,
      std::size_t Rounds = VSMC_RNG_THREEFRY_ROUNDS,
      typename Constants = ThreefryConstants<T, K>>
  class ThreefryGenerator;
\end{Verbatim}
The corresponding \rng engine is,
\begin{Verbatim}
  template <typename ResultType, typename T = ResultType,
      std::size_t K = VSMC_RNG_THREEFRY_VECTOR_LENGTH,
      std::size_t Rounds = VSMC_RNG_THREEFRY_ROUNDS,
      typename Constants = ThreefryConstants<T, K>>
  using ThreefryEngine =
      CounterEngine<ResultType, ThreefryGenerator<T, K, Rounds, Constants>>;
\end{Verbatim}
The default vector length and the number of rounds can be changed by
configuration macros listed in Table~\ref{tab:Configuration macros for
  counter-based RNG}. Similar to the implementation of the Philox algorithm,
there is no limit on the template parameter \verb|K| or \verb|Rounds| as long
as a suitable trait class \verb|Constant| is provided. The library provides
default constants for 64-bits unsigned integer type \verb|T| and \verb|K|
taking the values \verb|4|, \verb|8| and \verb|16|, taken from the
skein\footnote{\url{http://www.skein-hash.info}} hash algorithm, for which the
Threefish algorithm was originally developed for. Defaults for 32-bits \verb|T|
or \verb|K| taking the value \verb|2| are also provided, taken from
\textcite{Salmon:2011um}. Type aliases for these configurations are defined for
convenience,
\begin{Verbatim}
  template <typename ResultType>
  using Threefry2x32Engine = ThreefryEngine<ResultType, std::uint32_t, 2>;

  template <typename ResultType>
  using Threefry4x32Engine = ThreefryEngine<ResultType, std::uint32_t, 4>;

  template <typename ResultType>
  using Threefry2x64Engine = ThreefryEngine<ResultType, std::uint64_t, 2>;

  template <typename ResultType>
  using Threefry4x64Engine = ThreefryEngine<ResultType, std::uint64_t, 4>;

  template <typename ResultType>
  using Threefry8x64Engine = ThreefryEngine<ResultType, std::uint64_t, 8>;

  template <typename ResultType>
  using Threefry16x64Engine = ThreefryEngine<ResultType, std::uint64_t, 16>;
\end{Verbatim}
Type aliases for 32- and 64-bits \verb|ResultType| are also defined, as listed
in Table~\ref{tab:Counter-based RNG}. The performance data is in
Table~\ref{tab:Performance of ThreefryEngine}.

\begin{table}
  \input{tab/rng_engine_threefry}%
  \caption{Performance of \protect\texttt{ThreefryEngine}}
  \label{tab:Performance of ThreefryEngine}
\end{table}

\subsection{\texttt{RNG} and \texttt{RNGMini}}
\label{sub:RNG and RNGMini}

Note that, not all \rng{}s implemented by the library is available on all
platforms. The library also defines two type aliases \verb|RNG| and
\verb|RNG_64|, which are one of the \rng{}s listed in
Table~\ref{tab:Counter-based RNG}. The preference is in the order listed in
Table~\ref{tab:Default RNG}. The user can define the configuration macro
\verb|VSMC_RNG_TYPE| to override the choice made by the library.

\begin{table}
  \begin{tabularx}{\textwidth}{LLL}
    \toprule
    Alias  & Class & Availability \\
    \midrule
    \verb|RNG|    & \verb|ARS|         & \verb|VSMC_HAS_AES_NI| \\
                  & \verb|Threefry|    & Always available       \\
    \verb|RNG_64| & \verb|ARS_64|      & \verb|VSMC_HAS_AES_NI| \\
                  & \verb|Threefry_64| & Always available       \\
    \bottomrule
  \end{tabularx}
  \caption{Default \protect\rng}
  \label{tab:Default RNG}
\end{table}

\subsection{Seeding counter-based RNG}
\label{sub:Seeding counter-based RNG}

The singleton class template \verb|SeedGenerator| can be used to generate
distinctive seeds sequentially. For example,
\begin{Verbatim}
  auto &seed = SeedGenerator<void, unsigned>::instance();
  RNG rng1(seed.get()); // Construct rng1
  RNG rng2(seed.get()); // Construct rng2 with another seed
\end{Verbatim}
The first argument to the template can be any type. For different types,
different instances of \verb|SeedGenerator| will be created. Thus, the seeds
generated by two generators, \verb|SeedGenerator<T1>| and
\verb|SeedGenerator<T2>|, will be independent. The second parameter is the type
of the seed values. It can be any unsigned integer type. Classes such as
\verb|Particle<T>| will use the generator of the following type,
\begin{Verbatim}
  using Seed = SeedGenerator<NullType, VSMC_SEED_RESULT_TYPE>;
\end{Verbatim}
where \verb|VSMC_SEED_RESULT_TYPE| is a configuration macro which is defined to
\verb|unsigned| by default.

One can save and set the seed generator using standard \cpp streams. For
example,
\begin{Verbatim}
  std::ifstream is("seed.txt");
  if (is)
      is >> Seed::instance();    // Read seed from a file
  else
      Seed::instance().set(101); // Set it manually
  is.close();
  // Using Seed
  std::ofstream os("seed.txt");
  os << Seed::instance();        // Write the seed to a file
  os.close();
\end{Verbatim}
This way, if the simulation program needs to be repeated multiple times, each
time it will use a different set of seeds. A single seed generator is enough
for a single program. However, it is more difficult to ensure that each
computing node has a distinctive set of seeds in a distributed system. A simple
solution is to use the \verb|modulo| method of \verb|SeedGenerator|. For
example,
\begin{Verbatim}
  Seed::instance().modulo(n, r);
\end{Verbatim}
where $n$ is the number of processes and $r$ is the rank of the current node.
After this call, all seeds generated will belong to the equivalent class $s
\equiv r \mod n$. Therefore, no two nodes will ever generate the same seeds.
Note that, the seeds generated are not random at all. For any deterministic
\rng{}s, the same seeds always produce identical streams. However, distinctive
seeds does not always lead to independent streams. This seed generator is only
suitable for counter-based \rng{}s.

\section{Non-deterministic RNG}
\label{sec:Non-deterministic RNG}

If the \rdrand instructions are supported, the library also implements three
\rng{}s, \verb|RDRAND16|, \verb|RDRAND32| and \verb|RDRAND64|. They output 16-,
32-, and 64-bits random integers, respectively. The \rdrand instruction may not
return a random integer at all. The \rng engine will keep trying until it
succeeds. One can limit the maximum number of trials by defining the
configuration macro \verb|VSMC_RNG_RDRAND_NTRIAL_MAX|. A value of zero, the
default, means the number of trials is unlimited. If it is a positive number,
and if after the specified number of trials no random integer is return by the
\rdrand instruction, zero is returned. The performance data is in
Table~\ref{tab:Performance of non-deterministic RNG}.

\begin{table}
  \input{tab/rng_engine_rdrand}%
  \caption{Performance of non-deterministic \protect\rng}
  \label{tab:Performance of non-deterministic RNG}
\end{table}

\section{MKL RNG}
\label{sec:MKL RNG}

The \mkl library provides some high performance \rng{}s. The library implements
a wrapper class \verb|MKLEngine| that makes them accessible as \cppoo engines.
They are listed in Table~\ref{tab:MKL RNG}. Note that, \mkl{} \rng{}s perform
the best when they are used to generate vectors of random numbers. These
wrappers use a buffer to store such vectors. And thus they have much larger
state space than usual \rng{}s. Each \rng engines output by default 32-bits
integers. Similar to the counter-based \rng{}s, 64-bits variants are also
defined. The performance data is in Table~\ref{tab:Performance of MKL RNG}.

\begin{table}
  \begin{tabularx}{\textwidth}{LL}
    \toprule
    Class & \mkl \brng \\
    \midrule
    \verb|MKL_MCG59|         & \verb|VSL_BRNG_MCG59|         \\
    \verb|MKL_MT19937|       & \verb|VSL_BRNG_MT19937|       \\
    \verb|MKL_MT2203|        & \verb|VSL_BRNG_MT2203|        \\
    \verb|MKL_SFMT19937|     & \verb|VSL_BRNG_SFMT19937|     \\
    \verb|MKL_NONDETERM|     & \verb|VSL_BRNG_NONDETERM|     \\
    \verb|MKL_ARS5|          & \verb|VSL_BRNG_ARS5|          \\
    \verb|MKL_PHILOX4X32X10| & \verb|VSL_BRNG_PHILOX4X32X10| \\
    \bottomrule
  \end{tabularx}
  \caption{\protect\mkl{} \protect\rng}
  \label{tab:MKL RNG}
\end{table}

\begin{table}
  \input{tab/rng_engine_mkl}%
  \caption{Performance of \protect\mkl{} \protect\rng}
  \label{tab:Performance of MKL RNG}
\end{table}

\section{Multiple RNG streams}
\label{sec:Multiple RNG streams}

Earlier in section~\ref{sub:Particle} we introduced that \verb|particle.rng(i)|
returns an independent \rng instance. This is actually done through a class
template called \verb|RNGSet|. Three of them are implemented in the library.
They all have the same interface,
\begin{Verbatim}
  RNGSet<RNG> rng_set(N); // A set of N RNGs
  rng_set.resize(n);      // Change the size of the set
  rng_set.seed();         // Seed each RNG in the set with Seed::instance()
  rng_set[i];             // Get a reference to the i-th RNG
\end{Verbatim}
The first implementation is \verb|RNGSetScalar|. As its name suggests, it is
only a wrapper of a single \rng. All calls to \verb|rng_set[i]| returns a
reference to the same \rng. It is only useful when an \verb|RNGSet| interface
is required while the thread-safety and other issues are not important.

The second implementation is \verb|RNGSetVector|. It is an array of \rng{}s
with length $N$. It has memory cost $\calO(N)$. Many of the counter-based
\rng{}s have small state size and thus for moderate $N$, this cost is not an
issue. The method calls \verb|rng_set[i]| and \verb|rng_set[j]| return
independent \rng{}s if $i \ne j$. This implementation has the advantage that
the behavior of an algorithm can be entirely deterministic even when the
scheduling of parallel execution is dynamic, since each sample has its own
\rng.

Last, if \tbb is available, there is a third implementation \verb|RNGSetTBB|,
which uses thread-local storage (\tls). It has much smaller memory footprint
than \verb|RNGSetVector| while maintains better thread-safety. The performance
impact of using \tls is minimal unless the computation at the calling site is
trivial. For example,
\begin{Verbatim}
  std::size_t eval_pre(SingleParticle<T> sp)
  {
      auto &rng = sp.rng();
      // using rng to initialize state
      // do some computation, likely far more costly than TLS
  }
\end{Verbatim}
The type alias \verb|RNGSet| is defined to be \verb|RNGSetTBB| if \tbb is
available, otherwise defined to be \verb|RNGSetVector|. It is used by the
\verb|Particle| class template. One can replace the type of \rng set used by
\verb|Particle<T>| with a member type of \verb|T|. For example,
\begin{Verbatim}
  class T
  {
      public:
      using rng_set_type = RNGSetScalar<RNG>;
  };
\end{Verbatim}
will replace the type of the \rng set contained in \verb|Particle<T>|. Below is
a more advanced example of replacing \verb|rng_set_type| when using OpenMP for
parallelization.
\begin{Verbatim}
  template <typename RNGType>
  class RNGSetOMP
  {
      public:
      RNGSetOMP(std::size_t) : rng_(/* maximum number of OpenMP threads */)
      {
          seed();
      }

      void resize(std::size_t) {}

      void seed() { Seed::instance()(rng_.size(), rng_.begin()); }

      RNGType &operator[](std::size_t)
      {
          return rng_[omp_get_thread_num()];
      }

      private:
      Vector<RNGType> rng_;
  };

  class T
  {
      public:
      using rng_set_type = RNGSetOMP<RNG>;
  };
\end{Verbatim}
In this example, only a small number of \rng{} engines are created and it is
(mostly) thread-safe. However, there are quite a few situations where this
class is not suitable, with serialized nested parallel region being a primary
one. The library does not provide the above implementation by default. There
are too many cases that it can be misused.

\section{Distributions}
\label{sec:Distributions}

The library provides implementations of some common distributions. Some of them
are the same as those in the standard library, with \verb|CamelCase| names. For
example, \verb|NormalDistribuiton| can be used as a drop-in replacement of
\verb|std::normal_distribuiton|. This includes all of the continuous
distributions defined in the standard library. As stated in
section~\ref{sec:Vectorized random number generating}, all the distributions
defined in the library support vectorized random number generating. In the
following sections we introduce each distributions included in the library.

\subsection{Uniform bits distribution}
\label{sub:Uniform bits distribution}

The class template,
\begin{Verbatim}
  template <typename UIntType>
  class UniformBitsDistribution;
\end{Verbatim}
is similar to the standard library's \verb|std::independent_bits_engine|,
except that it always generates full size random integers. That is, let $W$ be
the number of bits of \verb|UIntType|, then the output is uniform on the set
$\{0,\dots,2^W - 1\}$. For example,
\begin{Verbatim}
  UniformBitsDistribution<std::uint32_t> rbits;
  rbits(rng); // Return 32-bits random integers
\end{Verbatim}
Let $\rmin$ and $\rmax$ be the minimum and maximum of the random integers
generated by \verb|rng|. Let $R = \rmax - \rmin + 1$. Let $r_i$ be consecutive
output of \verb|rng()|. If there exists an integer $M > 0$ such that $R = 2^M$,
then the result is,
\begin{equation*}
  U = \sum_{k = 0}^{K - 1} (r_k - \rmin) 2^{kM} \bmod 2^W
\end{equation*}
where $K = \Ceil{W / M}$. Unlike \verb|std::independent_bits_engine|, the
calculation can be vectorized, which leads to better performance. Note that,
all constants in the algorithm are computed at compile-time and the summation
is fully unrolled, and thus there is no runtime overhead. In the case $\rmin =
0$ and $M = W$, most optimizing compilers shall be able to generate
instructions such that the distribution does exactly nothing and returns the
results of \verb|rng()| directly. If there does not exist an integer $M > 0$
such that $R = 2^M$, then \verb|std::indepdent_bits_engine| will be used.

\subsection{Standard uniform distribution}
\label{sub:Standard uniform distribution}

All continuous distributions are built upon the standard uniform distribution.
And thus the performance and quality of the algorithm transferring random
integers to random floating point numbers on the set $[0, 1]$ are of critical
importance. The library provides five distributions, listed in
Table~\ref{tab:Standard uniform distributions}. They are all class template
with a single template type parameter \verb|RealType|, which is the floating
point result type. For each distribution, the random integers produced by
\rng{}s are transferred to 32- or 64-bits intermediate random integers through
\verb|UniformBitsDistribution| before they are further converted to floating
numbers. The integer type depends on \verb|RealType|, the range of the integers
produced by the \rng{}, $R$, and the configuration macros
\verb|VSMC_RNG_U01_USE_64BITS_DOUBLE|. The exact relations are listed in
Table~\ref{tab:Intermediate integer types of standard uniform distributions}.
In the remaining of this section, let $W$ be the number of bits of the
intermediate random integers, and $M$ be the number of significant bits
(including the implicit one) of \verb|RealType|. We also denote the input
random integers as $U$ and the output random real numbers as $X$.

\begin{table}
  \begin{tabularx}{\textwidth}{LL}
    \toprule
    Distribution & Support \\
    \midrule
    \verb|U01CCDistribution| & $[0, 1]$ \\
    \verb|U01CODistribution| & $[0, 1)$ \\
    \verb|U01OCDistribution| & $(0, 1]$ \\
    \verb|U01OODistribution| & $(0, 1)$ \\
    \verb|U01Distribution|   & $[0, 1)$ \\
    \bottomrule
  \end{tabularx}
  \caption{Standard uniform distributions}
  \label{tab:Standard uniform distributions}
\end{table}

\begin{table}
  \begin{tabularx}{\textwidth}{LlL}
    \toprule
    \verb|RealType| & Conditions & Integer type \\
    \midrule
    \verb|float| & $\log_2 R \ge 64$ & \verb|std::uint64_t| \\
                 & Otherwise         & \verb|std::uint32_t| \\
    \verb|double| & $\log_2 R \ge 64$ & \verb|std::uint64_t| \\
    & \verb|VSMC_RNG_U01_USE_64BITS_DOUBLE| & \verb|std::uint64_t| \\
    & Otherwise & \verb|std::uint32_t| \\
    \verb|long double| & Always & \verb|std::uint64_t| \\
    \bottomrule
  \end{tabularx}
  \caption{Intermediate integer types of standard uniform distributions}
  \label{tab:Intermediate integer types of standard uniform distributions}
\end{table}

\paragraph{\texttt{U01CCDistribution}}

This distribution produce random real numbers on $[0, 1]$, with the lower and
upper bounds inclusive. The specific algorithm is as the following,
\begin{align*}
  P &= \min\{W - 1, M\} \\
  V &= \begin{cases}
    U &\text{if } P + 1 < W \\
    \Floor{(U \bmod 2^{W - 1}) / 2^{W - P -2}} &\text{otherwise}
  \end{cases} \\
  Z &= (V \bmod 1) + V \\
  X &= 2^{-(P + 1)} Z
\end{align*}
The minimum and maximum are $0$ and $1$, respectively.

\paragraph{\texttt{U01CODistribution}}

This distribution produce random real numbers on $[0, 1)$, with the lower bound
inclusive and the upper bound never produced. The specific algorithm is as the
following,
\begin{align*}
  P &= \min\{W, M\} \\
  V &= \Floor{U / 2^{W - P}} \\
  X &= 2^{-P} V
\end{align*}
The minimum and maximum are $0$ and $1 - 2^{-P}$, respectively.

\paragraph{\texttt{U01OCDistribution}}

This distribution produce random real numbers on $(0, 1]$, with the upper bound
inclusive and the lower bound never produced. The specific algorithm is as the
following,
\begin{align*}
  P &= \min\{W, M\} \\
  V &= \Floor{U / 2^{W - P}} \\
  X &= 2^{-P} V + 2^{-P}
\end{align*}
The minimum and maximum are $2^{-P}$ and $1$, respectively.

\paragraph{\texttt{U01OODistribution}}

This distribution produce random real numbers on $(0, 1)$, with the lower and
upper bounds never produced. The specific algorithm is as the following,
\begin{align*}
  P &= \min\{W + 1, M\} \\
  V &= \Floor{U / 2^{W + 1 - P}} \\
  X &= 2^{-(P - 1)} V + 2^{-P}
\end{align*}
The minimum and maximum are $2^{-P}$ and $1 - 2^{-P}$, respectively.

\paragraph{\texttt{U01Distribution}}

It is now clear that the above four distributions actually produce ``fixed
point'' instead of ``floating point'' numbers. The output $X$ can be
represented exactly by the target \verb|RealType|. They have two advantages.
First, when it is important that the lower or upper bound is never produced, to
avoid underflow, overflow or other undefined behaviors in subsequent
calculations, they provide such assurance. Second, they usually can be executed
with only a couple of instructions by modern processors. And thus can have
better performance.

The main drawback is accuracy. If \verb|RealType| is \verb|float| or
\verb|long double|, then the difference is minimal, since the intermediate
random integers have more bits than the significant of the target floating
point type. The situation is a bit more tricky in the case of \verb|double| and
the intermediate random integers are 32-bits. In this case,
\verb|U01CODistribution| can only produce $2^{32}$ distinctive values while
\verb|double| can represent much more values exactly within the range $[0, 1)$.
In contrast, the standard library will use at least 53 random bits. This will
not matter in most realistic applications. In fact, random numbers produced by
\verb|U01CODistribution| passes all tests in the {\lnfigures\tbfigures
  TestU01}%
\footnote{\url{http://www.iro.umontreal.ca/~simardr/testu01/tu01.html}} library
that \verb|std::uniform_real_distribution| would pass, for a good \rng. In
other words, the quality of the \rng is the dominating factor.

However, there are situations where one do want the extra precision. For
example, the library implement the Normal distribution using the standard
Box-Muller method \parencite{Box:1958hv}, for performance consideration. Better
accuracy at the tail can only be archived by using procedures that can produce
values closer to zero than $2^{-32}$. In this case, there are two solutions.
The first is to define the configuration macro
\verb|VSMC_RNG_U01_USE_FIXED_POINT| to zero, and thus \verb|U01Distribution| is
no longer the same as \verb|U01CODistribution|. Instead, it behaves similarly
to \verb|std::generate_canonical|. More specifically,
\begin{align*}
  P &= \Floor{(W + M - 1) / W} \\
  K &= \max\{1, P\} \\
  X &= \sum_{k=0}^{K - 1} U_k 2^{-(K - k)W}
\end{align*}
The other solution is to define the configuration macro
\verb|VSMC_RNG_U01_USE_64BITS_DOUBLE| to a non-zero value, such that the
intermediate random integers will always be 64-bits for \verb|double| output.
This configuration macro also affects all other four distributions discussed
earlier.

\subsection{Distributions using the inverse method}
\label{sub:Distributions using the inverse method}

Table~\ref{tab:Distributions using the inverse method} lists all the
distributions implemented with the inverse method. The performance data of
these distributions is in Table~\ref{tab:Performance of distributions using the
  inverse method}. Note that, in this and other tables in the remaining of this
section, we shortened the class name for brevity. For example, in the table the
Cauchy distribution is listed as \verb|Cauchy| while the full declaration of
the class template is,
\begin{Verbatim}
  template <typename RealType>
  class CauchyDistribution;
\end{Verbatim}

\begin{table}
  \begin{tabularx}{\textwidth}{lllL}
    \toprule
    Distribution & Parameters & Support & \cdf \\
    \midrule
    \verb|Cauchy| & \verb|a|, \verb|b| & $(-\infty,\infty)$ &
    $\frac{1}{\pi}\arctan\Round[Big]{\frac{x - a}{b}} + \frac{1}{2}$ \\
    \verb|Exponential| & \verb|lambda| & $[0,\infty)$ &
    $1 - \EE^{-\lambda x}$ \\
    \verb|ExtremeValue| &\verb|a|, \verb|b| & $(-\infty,\infty)$ &
    $\exp\Curly[Big]{-\exp\Round[Big]{-\frac{x - a}{b}}}$ \\
    \verb|Laplace| & \verb|a|, \verb|b| & $(-\infty,\infty)$ &
    $\frac{1}{2} + \frac{1}{2}\mathrm{sgn}(x - a)\Round[Big]{1 -
      \exp\Curly[Big]{-\frac{\Abs{x - a}}{b}}}$ \\
    \verb|Logistic| & \verb|a|, \verb|b| & $(-\infty,\infty)$ &
    $\Round[Big]{1 + \exp\Curly[Big]{-\frac{x - a}{b}}}^{-1}$ \\
    \verb|Pareto| & \verb|a|, \verb|b| & $[a, \infty)$ &
    $1 - \Round[Big]{\frac{b}{x}}^a$ \\
    \verb|Rayleigh| & \verb|sigma| & $[0, \infty)$ &
    $1 - \exp\Curly[Big]{-\frac{x^2}{2\sigma^2}}$ \\
    \verb|UniformReal| & \verb|a|, \verb|b| & $[a, b)$ &
    $\frac{x - a}{b - a}$ \\
    \verb|Weibull| & \verb|a|, \verb|b| & $[0, \infty)$ &
    $1 - \exp\Curly[Big]{-\Round[Big]{\frac{x}{b}}^a}$ \\
    \bottomrule
  \end{tabularx}
  \caption{Distributions using the inverse method}
  \label{tab:Distributions using the inverse method}
\end{table}

\begin{table}
  \input{tab/rng_distribution_inverse}%
  \caption{Performance of distributions using the inverse method}
  \label{tab:Performance of distributions using the inverse method}
\end{table}

\subsection{Normal and related distribution}
\label{sub:Normal and related distribuiton}

The class template
\begin{Verbatim}
  template <typename RealType>
  class NormalDistribution;
\end{Verbatim}
implements the Normal distribution with \pdf,
\begin{equation*}
  f(x) =
  \frac{1}{\sqrt{2\pi\sigma^2}}\exp\Curly[Big]{-\frac{(x-\mu)^2}{2\sigma^2}},
\end{equation*}
using the Box-Muller method \parencite{Box:1958hv}. It is also used to
implement the Log-Normal distribution, $X = \EE^{m + s Z}$, where $Z$ follows
the standard Normal distribution,
\begin{Verbatim}
  template <typename RealType>
  class LognormalDistribution;
\end{Verbatim}
and the Levy distribution, $X = a + b / Z^2$,
\begin{Verbatim}
  template <typename RealType>
  class LevyDistribution;
\end{Verbatim}
The performance data of these distributions is in Table~\ref{tab:Performance of
  Normal and related distributions}.

\begin{table}
  \input{tab/rng_distribution_normal}%
  \caption{Performance of Normal and related distributions}
  \label{tab:Performance of Normal and related distributions}
\end{table}

\paragraph{Multivariate Normal distribution}

The library also implements the multivariate Normal distribution,
\begin{Verbatim}
  template <typename RealType, std::size_t Dim>
  class NormalMVDistribution;
\end{Verbatim}
If \verb|Dim| is zero (\verb|Dynamic|), then the distribution can only be
constructed with,
\begin{Verbatim}
  explicit NormalMVDistribution(std::size_t dim,
      const result_type *mean = nullptr, const result_type *chol = nullptr);
\end{Verbatim}
If \verb|Dim| is a positive integer, it can only be constructed with,
\begin{Verbatim}
  explicit NormalMVDistribution(
      const result_type *mean = nullptr, const result_type *chol = nullptr);
\end{Verbatim}
In either case, the parameter \verb|mean| is the mean vector. If it is a null
pointer, then it is assumed to be a zero vector. The parameter \verb|chol| is a
$d(d + 1)/2$-vector, where $d$ is the dimension. The vector is the lower
triangular elements of the Cholesky decomposition of the covariance matrix,
packed row by row. Libraries such as \lapack has routines to compute such a
matrix. Alternatively, one can use the covariance functionalities in the
library. See section~\ref{sec:Sample covariance}, which also provides a
concrete example of using the multivariate Normal distribution.

\subsection{Gamma and related distribution}
\label{sub:Gamma and related distribution}

The class template
\begin{Verbatim}
  template <typename RealType>
  class GammaDistribution;
\end{Verbatim}
implements the Gamma distribution with \pdf,
\begin{equation*}
  f(x) = \frac{\EE^{-x/\beta}}{\Gamma(\alpha)}\beta^{-\alpha}x^{\alpha-1}.
\end{equation*}
The specific algorithm used depends on the parameters. If $\alpha = 1$, it
becomes the exponential distribution. If $0 < \alpha < 0.6$, it is generated
through transformation of exponential power distribution
\parencite[sec~2.6]{Devroye:1986gi}. If $0.6\le\alpha<1$, then rejection method
from the Weibull distribution is used \parencite[sec.~3.4]{Devroye:1986gi}. If
$\alpha > 1$, then the method in \textcite{Marsaglia:2000vq} is used. There are
three related distributions,
\begin{Verbatim}
  template <typename RealType>
  class ChiSquaredDistribution;

  template <typename RealType>
  class FisherFDistribution;

  template <typename RealType>
  class StudentTDistribution;
\end{Verbatim}
They implement the $\chi^2$-distribution, the Fisher's $F$-distribution, and
the Student's $t$-distribution, respectively. The performance data of these
distributions, for different parameters are in Tables~\ref{tab:Performance of
  Gamma distribution} to~\ref{tab:Performance of Student's t-distribution}.

\begin{table}
  \input{tab/rng_distribution_gamma}%
  \caption{Performance of Gamma distribution}
  \label{tab:Performance of Gamma distribution}
\end{table}

\begin{table}
  \input{tab/rng_distribution_chisquared}%
  \caption{Performance of $\chi^2$ distribution}
  \label{tab:Performance of chi-squared distribution}
\end{table}

\begin{table}
  \input{tab/rng_distribution_fisherf}%
  \caption{Performance of Fisher's $F$-distribution}
  \label{tab:Performance of Fisher's F-distribution}
\end{table}

\begin{table}
  \input{tab/rng_distribution_studentt}%
  \caption{Performance of Student's $t$-distribution}
  \label{tab:Performance of Student's t-distribution}
\end{table}

\subsection{Beta distribution}
\label{sub:Beta distribution}

The class template
\begin{Verbatim}
  template <typename RealType>
  class BetaDistribution;
\end{Verbatim}
implements the Beta distribution with \pdf,
\begin{equation*}
  f(x) = \frac{\Gamma(\alpha + \beta)}{\Gamma(\alpha)\Gamma(\beta)}
  x^{\alpha - 1}(1 - x)^{\beta - 1}
\end{equation*}
The specific algorithm used depends on the parameters. If $\alpha = 1/2$ and
$\beta = 1/2$, or $\alpha = 1$ or $\beta = 1$, then the inverse method is used.
If $\alpha > 1$ and $\beta > 1$, the method in \textcite{Cheng:1978jl} is used.
Otherwise, let $K = 0.852$, $C = -0.956$, and $D = \beta + K\alpha^2 + C$. If
$\alpha < 1$, $\beta < 1$ and $D \le 0$, then Jöhnk's method
\parencite[sec.~3.5]{Devroye:1986gi} is used. In all other cases, one of the
switching algorithms in \textcite{Atkinson:1979es} is used. Note that, there is
no vectorized implementation at the moment for the switching algorithms. In
other cases, the vectorized generating shall provide considerable speedup. The
performance data is in Table~\ref{tab:Performance of Beta distribution}

\begin{table}
  \input{tab/rng_distribution_beta}%
  \caption{Performance of Beta distribution}
  \label{tab:Performance of Beta distribution}
\end{table}
